\chapter{Acknowledgments}

This thesis is the culmination of almost four years of work at the University of
Alberta and could not have been written without the help and support of my
mentors, coworkers, friends, and family.

I would like to thank my colleague and office mate Christopher Polachic for
tirelessly enduring my monologues on quantum-dot cellular automata, bouncing off
my ideas, and inspiration. Thanks to him I am equipped with a better
understanding of DMFT and the dynamic Hubbard model, and of the Canadian world
view. I am grateful for discussions on the books of Albert Camus.

My coworker Jin Xu has taught me about most things between spin systems and job
hunting, and given me some insight into Chinese culture, and I would like to
thank him for that. I am indebted to Roshan Achal for distractful office lunch
breaks and a fun year with the graduate physics student association. I would
like to thank Carl Chandler for writing a small software library together and
for tolerating my snoring in Denver. I owe a great trip to Boston and New York,
a missed flight, and a great time in Dresden to Fabian Zschocke, with whom I
shared the office for one year.

Marco Taucer and the whole group of Robert Wolkow have provided valuable
feedback to my ideas and calculations, and patiently explained their dangling
bonds and their understanding of and vision for quantum-dot cellular
automata. I also thank them for the experiments that really originally inspired
this work.

I would like to note the advise and support I have received from my supervisory
committee members, Mark Freeman, Massimo Bnonsegni, and Frank Marsiglio. I am
very grateful for them accepting and working with my tight deadlines. I also
thank Frank Marsiglio for a fantastic superconductivity course (and some
sleepless nights over assignments).

I especially thank my supervisor Kevin Beach. He has provided guidance and help
over the years, and an all-in-all very comfortable and enjoyable work
environment / relationship. I am grateful for my last supervisor Fakher Assaad
for giving me the idea to work with Kevin Beach and move to Edmonton, this nice
city with an underwhelming reputation in the middle of nowhere in western Canada.
% Kevin: last minute corrections

I particularly want to thank my girlfriend for encouraging me to keep going,
convincing me that this thesis can actually be finished, and for accepting me
working odd hours through the night. I thank her for always being there and
moving around half the world with me.

I thank Alberta Innovates - Technology Futures for paying my living for the
last two years. I am grateful for Kevin's NSERC Discovery grant for covering
travelling and other expenses.



% Christopher Polachic
% Jin Xu
% Carl Chandler
% Fabian Zschocke
% Piyush Jain
% Roshan Achal
% Marco Taucer, Lucian Livadaru, Robert Wolkow
% Kevin Beach
% Esly, family
%
% Fakher Assaad for sending me to Canada (?)
%
% Dominik, Jens, Christoph (?)
%
% Supervisory committee: Mark Freeman, Frank Marsiglio, Massimo Boninsegni

% Alberta Innovates - Technology Futures
% NSERC Discovery
